Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-09-02 submission/manuscript.Rmd
editing discussion

# Discussion
In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity.
We first characterized the misexpression of tissue-enriched genes within the _Kdm5c_-KO brain and identified substantial dysregulation of testis, liver, muscle, and ovary-enriched genes. Testis genes significantly enriched within the _Kdm5c_-KO amygdala and hippocampus are specific to germ cells and not expressed within testicular somatic cells. _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly express key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the _Kdm5c_-KO brain primarily expresses genes important for late spermatogenesis. We demonstrated that while _Kdm5c_ mutant sex did not influence whether sperm or egg-specific genes were misexpressed, female EpiLCs are more sensitive to germline gene de-repression. Germline genes can become aberrantly expressed in _Kdm5c_-KO cells via an indirect mechanism, as KDM5C is only bound to a subset of germline-enriched DEGs. Finally, we found KDM5C is dynamically regulated during ESC to EpiLC differentiation and promotes long-term germline gene silencing through DNA methylation at CpG islands. Therefore, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary. By systematically characterizing KDM5C's role in germline gene repression, including its interaction with known silencing mechanisms, we unveiled unique repressive mechansisms governing distinct classes of germline gene in somatic lineages. Furthermore, these data provide molecular footholds that can then be exploited to test the ultimate contribution of ectopic germline gene expression upon neurodevelopment.

It is important to note that while we highlighted KDM5C's regulation of germline genes, some germline-enriched genes are also expressed at the 2-cell stage and in na&#239;ve ESCs for their role in pluripotency and self-renewal. For example, _Dazl_ is primarily known for committing PGCs to the germline fate and regulating the translation of germline-specific mRNAs, but is also expressed at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], in na&#239;ve ESCs[@wellingDAZLRegulatesTet12015], and in the inner cell mass [@wellingDAZLRegulatesTet12015]. Based on the de-repression of _Dazl_ and _Zscan4c_ in _Kdm5c_-KO ESCs, KDM5C was thought to promote the 2-cell-to-ESC transition[[@guptaGenomewideScreenReveals2023]. Although expressed in na&#239;ve ESCs, _Dazl_ and other "self-renewal" germline genes are silenced during ESC differentiation into epiblast stem cells/EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021]. We found that while _Kdm5c_-KO EpiLCs also expressed _Dazl_, they did not express 2-cell specific genes. Misexpression of many germline genes in _Kdm5c_-KO EpiLCs may indicate they are differentiating into primordial germ cell-like cells (PGCLCs)[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014; @hayashiReconstitutionMouseGerm2011]. Yet, _Kdm5c_-KO EpiLCs had normal cellular morphology and properly expressed markers for primed pluripotency, including _Otx2_ which is known to repress EpiLC differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018]. Altogether, these data suggest _Kdm5c_-KO germline gene misexpression occurs ectopically in conjunction with typical developmental programs and the 2-cell-like state observed in _Kdm5c_-KO ESCs likely reflect KDM5C's primary role in germline gene repression.

Although eggs and sperm employ the same transcriptional programs for shared functions, e.g. PGC formation,  meiosis, and genome defense, some germline genes are sex specific. We found _Kdm5c_ mutant males and females expressed both sperm and egg-biased genes, indicating the mechanism of derepression is independent of a given germline gene's sex. However, organismal sex did greatly influence the degree of germline gene dysregulation, as female _Kdm5c_-KO EpiLCs had over double the number of germline-enriched DEGs compared to males. The lack of X-linked gene enrichment in females suggests that this greater sensitivity to germline gene misexpress is not due to XCI defects previously reported in _Kdm5c_-KO females[@samantaActivationXistEvolutionarily2022]. Sex differences in germline gene suppression may be a consequence of females having a higher dose of KDM5C than males, due to its escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008]. Intriguingly, females with heterozygous loss of _Kdm5c_ also had over double the number of germline DEGs than hemizygous knockout males, even though their level of KDM5C should be roughly equivalent to that of wild-type males. Altogether, these results suggests germline gene silencing mechanims differ between males and females, which warrants further study to identify the biological implications and underlying mechanisms.

Emerging work indicates many histone-modifying enzymes have non-cataltyic functions that influnce gene expression, sometimes even more potently than their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020]. KDM5C's catalytic activity may not be required for germline gene silencing, as it was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023].

@@

- Altogether, our results indicates a given chromatin modifier's catalytic and non-catlytic regulation of gene expression can change depending upon the locus and developmental stage.

Although the classic model for germline gene silencing is accumulation of CpGme at CGIs within their promoter. We found less than 30% of germline-enriched genes have CGIs, yet the majority of CGI-free germline genes still gained significant CpGme around their TSS.

- CGIs highly determined KDM5C recruitment to germline gene promoters.

- KDM5C loss also impacted CpGme at some KDM5C-unbound, CGI-free germline genes. Unclear what the repressive mechanism is for these genes.

In somatic cells, germline genes are highly methylated at promoter CGIs, which are typically unmethylated for other classes of genes.

- Germline gene suppression has focused on the accumulation of CpGme at CGIs

- DNAme and CpG islands

@@

- CGIs highly determine KDM5C recruitment to germline gene promoters.
			- KDM5C is known to be enriched at CGIs (in neurons? are these methylated? or is its germline CGI function different from its somatic CGI function?). 
		- Other studies on germline gene repressors have shown they are important for CGIme, unclear if they participate in non-CGI TSS CpGme
	- Altogether demonstrates different classes of germline genes use distinct silencing mechanisms

It is yet unknown how KDM5C is recruited to germline gene promoters, given that KDM5C itself does not contain domains for sequence-specific binding[@iwaseXLinkedMentalRetardation2007]. In HeLa cells and ESCs[@tahilianiHistoneH3K4Demethylase2007; @guptaGenomewideScreenReveals2023], KDM5C assocaiates with members of the polycomb repressive complex 1.6 (PRC1.6), which is recruited to germline gene promoters through E2F6/DP1 and MGA/MAX heterodimers[@endohPCGF6PRC1SuppressesPremature2017; @stielowMGAL3MBTL2E2F62018]. While MAX and E2F6 motifs were enriched at KDM5C-bound germline genes in EpiLCs, only about one-third of promoters contained their consesus sequence. Additionally, some germline genes can become active in _Kdm5c_-KO cells independent of KDM5C binding to their promoters. Germline genes unbound by KDM5C were primarily late-stage spermatogenesis genes and significantly enriched for RFX transcription factors, including RFX2, a central regulator of spermiogenesis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016]. Another notable KDM5C-unbound DEG is _Stra8_, a meiotic transcription factor that is turned on in germ cells with retionic acid signaling and DAZL expression[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006]. Misexpression of _Dazl_ and _Rfx2_ and their downstream targets in _Kdm5c_-KO cells suggests that once activated, ectopic germline genes can then turn on other aberrant germline programs to loosely mimic germ cell development.

@@

## Immunocytochemistry (ICC)
ICC of DAZL in EpiLCs was performed by first growing cells on fibronectin-coated coverslips. Cells were then washed thrice with phosphobuffered saline (PBS), fixed in 4% paraformaldehyde, washed thrice in PBS, and blacked in PBS containing 0.3% Triton X-100, and 5% fetal bovine serum for 1 hour. They were then washed thrice with PBS and incubating in primary antibody (Rabbit anti DAZL, abcam ab34139, 1:200) in the blocking solution overnight at 4°C with gentle rocking. The next day, cells were rinsed thrice with PBS, and incubated in secondary antibody (Alexafluor 488 Invitrogen #710369, 1:1,000) with DAPI (1:1,000) in blocking buffer for 1 hour at room temperature. Coverslips were then rinsed thrice in PBS, and mounted onto slides using Prolong Gold (Invitrogen #P36930). Images were taken blinded for genotype, chosen based on similar levels of DAPI signal, and quantified via ImageJ before unblinding.

>>>>

# Discussion
In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity. We first characterized tissue-enriched genes expressed within the _Kdm5c_-KO brain and identified substantial dysregulation of testis, liver, muscle, and ovary-enriched genes. Testis genes significantly enriched within the _Kdm5c_-KO amygdala and hippocampus are specific to the germline and not expressed within somatic cells. _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly express key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the adult brain primarily expresses genes important for late spermatogenesis. We demonstrated that although _Kdm5c_ mutant sex did not influence whether sperm or egg-specific genes were misexpressed, female EpiLCs are more sensitive to germline gene de-repression. Germline genes can become aberrantly expressed in _Kdm5c_-KO cells via an indirect mechanism, such as activation via ectopic RFX transcription factors. Finally, we found KDM5C is dynamically regulated during ESC to EpiLC differentiation to promote long-term germline gene silencing through DNA methylation at CpG islands. Therefore, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary. By systematically characterizing KDM5C's role in germline gene repression, including its interaction with known silencing mechanisms, we unveiled repressive mechansisms governing distinct classes of germline gene in somatic lineages. Furthermore, these data provide molecular footholds which can be exploited to test the overarching contribution of ectopic germline gene expression upon neurodevelopment.

**make edits**
Although eggs and sperm employ the same transcriptional programs for shared functions, e.g. PGC formation,  meiosis, and genome defense, some germline genes are sex specific. We found _Kdm5c_ mutant males and females expressed both sperm and egg-biased genes, indicating the mechanism of derepression is independent of a given germline gene's sex. However, organismal sex did greatly influence the degree of germline gene dysregulation, as female _Kdm5c_-KO EpiLCs had over double the number of germline-enriched DEGs compared to males. The lack of X-linked gene enrichment in females suggests that this greater sensitivity to germline gene misexpress is not due to XCI defects previously reported in _Kdm5c_-KO females[@samantaActivationXistEvolutionarily2022]. Sex differences in germline gene suppression may be a consequence of females having a higher dose of KDM5C than males, due to its escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008]. Intriguingly, females with heterozygous loss of _Kdm5c_ also had over double the number of germline DEGs than hemizygous knockout males, even though their level of KDM5C should be roughly equivalent to that of wild-type males. Altogether, these results suggests germline gene silencing mechanims differ between males and females, which warrants further study to identify the biological implications and underlying mechanisms.

It is important to note that while we highlighted KDM5C's regulation of germline genes, some germline-enriched genes are also expressed at the 2-cell stage and in na&#239;ve ESCs for their role in pluripotency and self-renewal[@suzukiLossMAXResults2016; @torres-padillaTransposonsTotipotency2020;@yangChemicalinducedChromatinRemodeling2022; @guptaGenomewideScreenReveals2023]. For example, while _Dazl_ is primarily known for regulating the translation of germline-specific mRNAs, it is also expressed at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], in na&#239;ve ESCs[@wellingDAZLRegulatesTet12015], and in the inner cell mass [@wellingDAZLRegulatesTet12015]. KDM5C was thought to promote the 2-cell-to-ESC transition, given _Dazl_ and _Zscan4c_ are de-repressed in _Kdm5c_-KO ESCs[@guptaGenomewideScreenReveals2023]. Although expressed in na&#239;ve ESCs, "self-renewal" germline genes like _Dazl_ are silenced during ESC differentiation into epiblast stem cells/EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021]. We found that while _Kdm5c_-KO EpiLCs also express _Dazl_, they did not express 2-cell-specific genes. These data suggest the 2-cell-like state observed in _Kdm5c_-KO ESCs likely reflects KDM5C's primary role in germline gene repression. Germline gene misexpression in _Kdm5c_-KO EpiLCs may indicate they are differentiating into primordial germ cell-like cells (PGCLCs), rather than de-differentiating into 2-cell-like cells[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014; @hayashiReconstitutionMouseGerm2011]. Yet, _Kdm5c_-KO EpiLCs had normal cellular morphology and properly expressed markers for primed pluripotency, including _Otx2_ which is known to repress EpiLC differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018]. In addition to unimpaired EpiLC differentiation, _Kdm5c_-KO gross brain morphology is overall normal[@iwaseMouseModelXlinked2016] and hardly any brain-specific genes were significantly dysregulated. Thus, ectopic germline gene expression occurs in parallel with proper _Kdm5c_-KO differentiation and somatic development.

- In somatic cells, germline genes are highly methylated at promoter CGIs, which are typically unmethylated for other classes of genes.
- The classic model for germline gene silencing is accumulation of CpGme at CGIs within their promoter, which is facilitated by E2F6/DP1 and MAX/MGA binding sites
- KDM5C does not have sequence specific binding, associates with PRC1.6 but likely not the major mechanisms of recruitment 
- KDM5C highly enriched at meiotic, promoter CGI germline genes
- A plausible mechanism for KDM5C-mediated germline gene silencing is KDM5C's removal of H3K4me2/3 at germline gene CGIs[@scandagliaLossKdm5cCauses2017; @aladhamiSystematicIdentificationFactors2023], given that (H3K4me2/3) oppose DNMTs and primarily do no colocalize with CpGme.
- Indeed we observed H3K4me2/3 accumulation at germline TSS and impaired CGI methylation at select germline genes. Loss of KDM5C strongly impairs CpGme at a subset of CGI-containing germline genes.
- However, Emerging work indicates many histone-modifying enzymes have non-cataltyic functions that influnce gene expression, sometimes even more potently than their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020].
- KDM5C's catalytic activity was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023]. In EpiLCs, Dazl still gained CpGme even with elevated H3K4me2. This suggests other redundant silencing mechanisms to ensure eventual CpGme of dazl, however other germline CGIs require KDM5C to ensure sufficient methylation
- Furthermore, our results indicates a histone modifier's catalytic and non-catlytic roles can change depending upon the locus and developmental stage.

- It is yet unknown how KDM5C is recruited to germline gene promoters, given that KDM5C itself does not contain domains for sequence-specific binding[@iwaseXLinkedMentalRetardation2007]. In HeLa cells and ESCs[@tahilianiHistoneH3K4Demethylase2007; @guptaGenomewideScreenReveals2023], KDM5C assocaiates with members of the polycomb repressive complex 1.6 (PRC1.6), which is recruited to germline gene promoters through E2F6/DP1 and MGA/MAX heterodimers[@endohPCGF6PRC1SuppressesPremature2017; @stielowMGAL3MBTL2E2F62018]. While MAX and E2F6 motifs were enriched at KDM5C-bound germline genes in EpiLCs, only about one-third of promoters contained their consesus sequence.

- Our list of germline genes
- We found less than 30% of germline-enriched genes have CGIs, yet the majority of CGI-free germline genes still gained significant CpGme around their TSS. 
	- CGI-containing = meiotic
- Loss of KDM5C binding at a subset of CGI-containing germline genes in EpiLCs, e.g. D1Pas1, Naa11, strongly impairs CpGme, resulting in their ectopic transcription in both EpiLCs and in the mature brain.
- Our germline gene list enabled us to identify germline-enriched genes activated independent of

- This analysis revealed some germline genes can become active in _Kdm5c_-KO cells independent of KDM5C binding to their promoters. 
- Germline genes unbound by KDM5C primarily late-stage spermatogenesis genes that lacked CGIs and significantly enriched for RFX transcription factors, including RFX2, a central regulator of spermiogenesis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016].

Another notable KDM5C-unbound DEG is _Stra8_, a meiotic transcription factor that is turned on in germ cells with retionic acid signaling and DAZL expression[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006]. Misexpression of _Dazl_ and _Rfx2_ and their downstream targets in _Kdm5c_-KO cells suggests that once activated, ectopic germline genes can then turn on other aberrant germline programs to loosely mimic germ cell development.

- KDM5C loss also activates RFX2 transctiption factors whose binding sites are enriched at non-CGI spermiogenesis genes.

- CGIs are resistant
	- https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5001583/
	- https://pubmed.ncbi.nlm.nih.gov/14504655/

@@

-

@@

- CGIs highly determine KDM5C recruitment to germline gene promoters.
		- Other studies on germline gene repressors have shown they are important for CGIme, unclear if they participate in non-CGI TSS CpGme

@@
